Occupation Gender Segregation: Empirical Evidence from a Matching Model with Transfers

### 2020-03-09 paper/Paper_2019_February_v2.tex
final version submitted to ReStud

\documentclass[12pt]{article}

@@

\usepackage{apacite}
\doublespacing

@@

\topmargin=-1in      
\evensidemargin=0in     
\oddsidemargin=0in      
\textwidth=6.5in        
\textheight=9.0in       
\headsep=0.25in

%\renewcommand{\baselinestretch}{2.0}

%\usepackage[margin = 1.5in]{geometry}

\newcommand*{\graphicPath}{"/Users/Miriam/OneDrive/Box Sync/TYPlocal/output/graphs/matlabwages/"}% 
\newcommand*{\TablePath}{"/Users/Miriam/OneDrive/Box Sync/TYPlocal/output/tables/"}% 
\graphicspath{{\graphicPath}}
\makeatletter
\def\input@path{{\TablePath}}
\makeatother

\title{Gender Segregation in Occupations: Empirical Evidence from a Matching Model with Transfers}
\author{Miriam Larson-Koester\footnote{Federal Trade Commission, mlarsonkoester@ftc.gov. The views expressed are those of the author and do not necessarily reflect those of the Federal Trade Commission. I am grateful to Francesca Molinari, Richard Mansfield and Victoria Prowse for their advice and support. For helpful discussions and insights I thank Francine Blau, Panle Barwick, Mallika Thomas, Larry Blume, JF Houde, Yang Zhang, Angela Cools, Jorgen Harris, Caroline Walker, and numerous seminar participants. All errors are mine.}}
%\affil[1]{Department of Economics, Cornell University}

@@

% with angela edits
% The sorting of men and women into different occupations is a persistent feature of the labor market and is increasingly important in explaining the gender wage gap in the U.S.

\abstract{Women remain concentrated in certain occupations despite the drastic increase in women's labor force participation in the U.S. since 1960. I examine whether occupations remain segregated because workers prefer to enter occupations that already employ more of their own gender. I build a model of labor supply and demand in which firms care about the gender and wages of their employees, and workers get utility from their occupation, wage, and the number of women in their occupation. Using a Bartik instrumental variables strategy, I find that women prefer to enter into female-dominated occupations, but men show no evidence of gender preference. In a world without women's gender preference, occupations would be 50 percent less segregated and the gender wage gap 18 percent smaller. However, women's preference for working with women would have to be almost twice as strong for past segregation to affect current segregation.}

% I built a model ... before us data labor market where workers have gender preferences and labor supply and demand... build a model of labor supply and demand where  gender prefrences affect occupational sorting

% in a world without gender pref, 50% less segregation and 18% less wage gap. wages adjust to compensating women for their preference. wage adjusment counteracts the preference

% gitkraken... sublime

%The sorting of men and women into different occupations is a persistent feature of the labor market and is increasingly important in explaining the gender wage gap in the U.S. \cite{Blau2017}. With U.S. data, I structurally estimate preferences over worker preferences over the gender in their occupation, and the reservation wage gap for men and women. Using a Bartik instrumental variables strategy, I find that women prefer to enter into female-dominated occupations, but men show no evidence of gender preference. Womens' preference for gender accounts for almost 50\% of  gender segregation. Wages adjust to compensate women for their preference, which reduces segregation, and explains 18\% of the gender wage gap. Simulations indicate that wage adjustment eliminates historical path dependence in segregation patterns.}

%The sorting of men and women into different occupations is a persistent feature of the labor market and is increasingly important in explaining the gender wage gap in the U.S. \cite{Blau2017}. Using U.S. data, I structurally estimate the willingness-to-pay for male and female workers by occupation, and the preferences of male and female workers for wages, exogenous occupation attributes, and occupation gender composition. Using a shift-share instrumental variables strategy, I find that women prefer to enter into more female occupations, and that this preference increases segregation in the cross-section and accentuates the impact of labor supply and demand shocks on gender composition over time. This gender preference is so strong that based on labor supply alone, most occupations have multiple equilibria, meaning that historical segregation could determine whether an occupation is male or female. However in the labor market, unlike in a classic Schelling model, compensating differentials moderate the impact of the gender preference, and ensure that all occupations have a unique equilibrium in fraction female. Therefore even at the high level of gender preference I estimate, manipulating the gender composition of an occupation alone has no long run impact on segregation.

@@

% LFP has skyrocketed but occupations still super segregated. policymakers care about this and why it is. we don't know. my question. why it's hard. incorporate lit review

Women's labor force participation has skyrocketed in the U.S. since 1960, but men and women still go into very different occupations. As of 2009, approximately 50\% of women would need change jobs in order to to achieve an equal number of men and women in every occupation, and this gap is unlikely to close soon \cite{Blau2013}. Recent literature has identified some contributing causes of segregation: preferences over amenities (\citeA{Olivieri2014a}, \citeA{Wiswall2013})\footnote{For overview see \citeA{ Bertrand2011} and \citeA{Cortes2017a}.}, productivity and skill differences (\citeA{Baker2015}), and occupational and educational barriers (\citeA{Hsieh2016}). I identify another possible cause of the persistence of segregation: workers valuing the gender composition of occupations.

In particular, this paper examines whether current segregation depends on past segregation because workers care about the gender of their occupation in addition to intrinsic characteristics. This is important because the right policy to address segregation, if any, will depend on what mechanisms are at play. Policymakers have expressed concern about occupational segregation, for example the lack of women in STEM and men in nursing. If these occupations are male or female for historical reasons, then segregation itself may be a barrier to individual careers and the efficiency of the economy as a whole. If, on the other hand, occupations are male or female based on intrinsic characteristics of workers or jobs, then interventions that seek to change the gender composition of occupations alone are likely inappropriate, and policymakers concerned about gender segregation should examine policies to address more specific causes as they deem necessary.

%Recent research using job satisfaction data suggests that women prefer not to work in male occupations \cite{Usui2008, Lordan2015}. Such a preference could operate at the occupation level, through gender identity effects \cite{Akerlof2000}, or through the work environment itself. Since the fraction female in an occupation evolves endogenously according to worker choices, preferences over the fraction female can produce rapid changes in gender composition at the occupation level. Rapid changes in composition can be referred to as ``tipping", and this was documented in a number of U.S. occupations from 1940-1980 by \citeA{Pan2010}.

%Tipping can result from two distinct mechanisms. First, group preferences can create a feedback loop that accelerates changes in composition caused by changes in the underlying attractiveness of a occupation, or labor demand. Second, if group preferences are strong enough, it may become profitable for an occupation to hire only men or women. This can mean that whether an occupation was historically male or female can affect its gender composition today.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%(ii) why is it hard to answer, 
%(iii) how does the paper answer the question, 
%(iv) what it novel about that method/data used to answer the question, and 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%(ii) why is it hard to answer, 
Identifying a preference over occupation fraction female is a difficult empirical problem that involves disentangling many other causes of gender segregation. Workers likely care about intrinsic characteristics of occupations, such as skills and job amenities (see for example \citeA{DeLeire2004} and \citeA{Reed1994}), as much as about the gender of their coworkers (see for example \citeA{Lordan2015} and \citeA{Usui2008}). At the same time, firms may care about which gender is perceived as more productive or valuable, as well as which gender of worker is cheaper to hire. Simply looking at which occupations contain more women will not allow me to distinguish firm and worker preferences, or worker preferences over the fraction female.

%(iii) how does the paper answer the question, 
The key innovation of this paper, building on the insights of \citeA{Choo2006}, \citeA{Salanie2014a}, and \citeA{Dupuy2017}, is that the distribution of wages can be used to separately identify worker and firm preferences. The intuition is that the right tail of wages tells us more about what firms are willing to pay for workers of a given type, while the left tail tells us more about what workers are willing to accept for a given type of job. Assuming that wage offers are distributed lognormally, and that firms do not care about the productivity of individuals within gender,\footnote{Unobserved productivity of individual workers could be included in theory, but would result in very weak identification and likely computational intractability because it would involve joint estimation of the model and integration over multiple sources of unobserved heterogeneity to interpret the wage distribution.} one can use Maximum Likelihood to estimate firm preferences and the wage offer distribution for male and female workers by occupation. Although the estimation strategy in this paper was developed specifically for segregation in the labor market, it could be used in any context in which a price mechanism clears a two-sided market.

The second identification challenge is to disentangle preferences for the number of women in an occupation from other occupation attributes. If an occupation is becoming more attractive to women, we will observe the fraction female in the occupation going up, and more women entering the occupation, and may incorrectly infer that women are entering the occupation because there are more women, a classic problem of omitted variable bias. To solve this problem I find variation in the fraction female by occupation that is plausibly exogenous to changes in other occupation amenities using Bartik-style instrumental variables. The main identifying assumption is that changes to occupations that workers care about are not correlated across occupations.

For tractability I treat occupation choice as a static choice. Workers choose occupation once at the beginning of their career, based on the contemporary characteristics of the occupations, including fraction female. Therefore, the fraction female in each occupation only updates across cohorts of workers.\footnote{Allowing workers to choose occupation myopically more than once during their lifetime, would lead to faster changes in occupation fraction females, but not otherwise change the dynamics of the model.} Lifetime wages are set to clear the market in static equilibrium for each cohort of men and women, which requires a large market assumption (continuum of workers and jobs of each type) to guarantee a unique wage equilibrium.

The data moments to identify the model are the shares of male and female workers by gender from the 1960-2000 U.S. Censuses and 2012 3-year ACS, and estimates of lifetime income by occupation, gender, and year constructed using income quantiles from the Census data combined with transition rates from the SIPP 2004 and 2008 panels. I take a two-step estimation approach, first estimating the firm side using maximum likelihood, then the worker side using instrumental variables regression, taking the wage offer distribution estimated on the firm side as data. The structure of the model allows me to simulate transition paths in the fraction female by cohort and determine there is path dependency, all while solving for equilibrium wages and fixing other firm and worker preferences.

%In the first stage of estimation, I find that firms are often willing to pay much more to hire male than female workers, but the degree to which this is true varies greatly across occupation. This gap may reflect discrimination, differences in labor force attachment over the lifetime, skill differences, or job sorting not captured in my occupation categories. The gap in the willingness to pay for female vs. male workers is lower in female dominated occupations. This might be because women are selecting into occupations in which they are more valued or have comparative advantage, or because the way work is valued might vary with the fraction female in the occupation \cite{Levanon2009, Harris2018}, or some combination of both.

@@

I find that women care strongly about the number of women in an occupation, but no evidence that men care about the number of women. The point estimate in my preferred specification is very high, with an occupation change from 25\% to 75\% female being equivalent to an extra \$3 million in lifetime income for a woman. The gender preference leads to more segregated outcomes. With no gender preference, the model predicts no occupations with fraction female greater than 70\% or less than 10\%, and a Duncan segregation index of 24\% in 2012, meaning 24\% of male or female workers would have to change occupations to make all occupations 50\% female. With the gender preference, the Duncan index is predicted to be almost twice as high at 47\%.\footnote{The Duncan index in my observed 2012 data is 41\%.}

% *cite tipping lit more thoroughly? CONSISTENT WITH THESE STYLIZED FACTS

Labor supply or demand shocks that affect the fraction female are reinforced by a feedback loop from the gender preference. An illustrative case study is insurance adjustors, which is one of many occupations observed to have moved from male to female by \citeA{Pan2010}. I find that more women began to become insurance adjusters because firms' demand for women in this occupation rose. Then as more women entered, the occupation became more attractive to women. This in turn made women cheaper to hire, which increased labor demand for female insurance adjusters, providing further reinforcement of the feminization of the occupation. My model predicts that if it were not for women preferring to work with women, insurance adjusters would have become only 50\% female, rather than the observed over 70\% female (from a starting point of 20\% female in 1960). Thus my estimated model provides one possible mechanism that is consistent with the stylized fact that occupations tend to move rapidly from male to female. This phenomenon is referred to as ``tipping" by \citeA{Pan2010}.\footnote{``Tipping" is also a documented phenomenon in racial segregation by neighborhoods \cite{Card2008} and racial composition of schools \cite{Caetano2017}.}

%If either men or women have preferences over the fraction female of occupations then we could observe multiple equilibria and ``tipping" patterns that result from moving between these equilibria. However, despite the extremely strong in-group preference that I estimate on the part of women, I find that equilibrium wages are able to adjust to prevent the emergence of multiple equilibria, and therefore tipping between equilibria. An even stronger preference would be required to produce these phenomena. I cannot rule out such a strong preference given the imprecision of my preferred specification, but such a strong preference seems economically implausible given the already large magnitude of my preferred specification.

% going from 25% to 75% female is equivalent to about $3 million in additional lifetime income
% going from 0% to 100% female is about equivalent to an extra 6 million dollars in lifetime income... yes...
% and yet I still don't see multiple equilibria????

@@

@@

Although the gender preference I estimate is very strong, it is largely mitigated by compensating differentials. In a model without endogenous wages, all occupations eventually converge to either 0\% or 100\% female. Wage adjustment leads to more moderate outcomes. As women enter an occupation, firms can hire the same number of women at lower cost, and lower wages dampen the increase in female labor supply. Likewise as women exit an occupation, firms must pay the remaining women more, which slows the exit of women. I find that these compensating differentials explain 18\% of the difference in lifetime income between men and women, with women in highly female dominated occupations suffering the largest loss in earnings. The result that as women enter an occupation, wages go down is consistent with previous literature \cite{Levanon2009, Harris2018}.

%For example, if it were not for womens' preference to work with women, women in the female-dominated fields of ``Health Service Occupations" and ``Health Assessment and Treating" would earn just as much as men.

@@

The presence of a gender preference could make it most profitable for occupations to hire only men or majority women. In this case, the segregation today may have been selected due to initial conditions such as historical barriers or norms \cite{Schelling1971, Pan2010}. However the gender preference I estimate would have to be about twice as strong to cause current occupation sorting to depend on past occupation sorting.

There are two important takeaways from this paper for policymakers. First, short run shocks to the fraction female in occupations, such as temporarily pushing more men into nursing or women into STEM, will not have long run consequences since there is no historical inertia to overcome. Policymakers seeking to reduce segregation should find specific supply and demand factors that concern them and address them directly, whether it be occupation amenities or discrimination. Second, policymakers should keep in mind that changes to labor supply and demand might have an outsized impact on segregation due to the feedback loop of women wanting to work with women. Making an occupation more attractive to women might have the unintended consequence of making the occupation highly female dominated, and thereby lowering the wages for those women over time through a compensating differential.

Lastly, future research should address the source of the gender preference to determine it is due to gender identity, or workplace environment and amenities. If workplace environment and amenities are the cause, then policymakers who care about reducing segregation and the gender wage gap might consider how to address these to make male-dominated occupations more welcoming to women.

%There are two policy implications from this result: First, policies that only temporarily change the number of men and women in an occupation have no impact in the long run. Because you can't shock policy. and because you can't care about segregation in and of itself, need to care about direct causes. Second, sorting patterns are not historically dependent. For example, if all occupations were 50\% female in 1960, gender segregation would still be very similar to what it is today.

@@

Section \ref{model} introduces the structure of the transferable utility matching model. Section \ref{equilibrium} describes equilibrium wages. Section \ref{empirical} describes the empirical specification and first stage of estimation, recovering the firm side parameters and reservation wages. Section \ref{stage2} describes the second stage of estimation, the decomposition of worker utility taking reservation wages as given. Section \ref{data} describes the data sources. Section \ref{results} presents results, and Section \ref{counterfactuals} presents simulations of segregation dynamics, including counterfactuals.

@@

%\section{Literature Review} \label{literature}

%A number of papers have studied worker preferences for tasks and amenities by occupation. \citeA{asadullah2008} and \citeA{Reed1994} find that men and women have similar preferences over amenities using job satisfaction surveys and job transitions respectively. On the other hand, \citeA{Goldin2011} and \citeA{Goldin2014} find higher demand from female workers for flexibility and shorter hours to be important in explaining sorting and wage differentials. I find large variation in male and female preferences for different occupations, which could include any occupation amenities that are fixed over time, such as flexibility, task requirements,\footnote{There is evidence that changes in task requirements over time is important in explaining gender gaps \cite{Black2010, atalay2017}. In a robustness check I control for the time-varying task measures of \citeA{atalay2017} and find qualitatively similar results for gender preference over fraction female.} hours, or risk.

%Several papers look for preference over the gender composition of the occupation. Using job satisfaction data from the NLSY, \citeA{Usui2008} finds that women dislike male occupations and receive a compensating differential to work in them. Similarly \citeA{Lordan2015} use job satisfaction and transition data to find that even controlling for occupation task measures and firm level fraction female, there is still evidence for women disliking male occupations.\footnote{The estimates in \citeA{Lordan2015} are robust even to controls for firm level fraction female and remain even after the inclusion of occupation task measures, suggestive of at least some preference over the gender composition at the occupation level.} Unlike this literature my paper explicitly models how such a preference factors into occupation choice, including the impact of compensating differentials on equilibrium wage levels.

%The methodological contribution of my paper builds on a long legacy of literature on theoretical matching. \citeA{Shapley1972a} showed that with transferable utility the stable matching coincides with the competitive equilibrium. \citeA{Crawford1981} and \citeA{RothSotomayor} further developed mechanisms for convergence to the competitive equilibrium, and conditions for the uniqueness of the prices, or wages, that support it. More recently \citeA{Choo2006} pioneered empirical estimation of transferable utility models with an application to the marriage market. \citeA{Salanie2014a} and \citeA{Dupuy2017} have presented the theoretical possibility of using observed transfers for identification. My paper is the first to my knowledge to take this insight to an empirical application, using wages to disentangle firm and worker preferences in the labor market.

%A possible consequence of workers having preferences over groups is ``tipping", or rapid movements in segregation patterns. The literature on tipping includes the seminal work of \citeA{Card2008}, identifying tipping points for racial segregation in neighborhoods, \citeA{Caetano2017} finding tipping and multiple equilibria in the racial composition of schools, and of course \citeA{Pan2010} using regression discontinuity to find tipping points in fraction female in occupations. I focus on disentangling the mechanisms underlying tipping, and am the first to my knowledge to examine the role of equilibrium wages or prices in such a market.

%My setting is different because of wages?

>>>>

\documentclass[11pt]{article}

@@

\usepackage{apacite}

\doublespacing

@@

\topmargin=-0.45in      
\evensidemargin=0in     
\oddsidemargin=0in      
\textwidth=6.5in        
\textheight=9.0in       
\headsep=0.25in

\newcommand*{\graphicPath}{"/Users/Miriam/Box Sync/TYPlocal/output/graphs/matlabwages/"}% 
\newcommand*{\TablePath}{"/Users/Miriam/Box Sync/TYPlocal/output/tables/"}% 
\graphicspath{{\graphicPath}}
\makeatletter
\def\input@path{{\TablePath}}
\makeatother

\title{Occupation Gender Segregation: Empirical Evidence from a Matching Model with Transfers}
\author{Miriam Larson-Koester\footnote{Federal Trade Commission, mlarsonkoester@ftc.gov. The views expressed are those of the author and do not necessarily reflect those of the Federal Trade Commission. I am grateful to Francesca Molinari, Richard Mansfield and Victoria Prowse for their advice and support. For helpful discussions and insights I thank Francine Blau, Panle Barwick, Mallika Thomas, Larry Blume, JF Houde, Yang Zhang, Angela Cools, Jorgen Harris, Caroline Walker, and numerous seminar participants. All errors are mine.}}
%\affil[1]{Department of Economics, Cornell University}

@@

% with angela edits

\abstract{The sorting of men and women into different occupations is a persistent feature of the labor market and is increasingly important in explaining the gender wage gap in the U.S. \cite{Blau2017}. Using U.S. data, I structurally estimate the willingness-to-pay for male and female workers by occupation, and the preferences of male and female workers for wages, exogenous occupation attributes, and occupation gender composition. Using a shift-share instrumental variables strategy, I find that women prefer to enter into more female occupations, and that this preference increases segregation in the cross-section and accentuates the impact of labor supply and demand shocks on gender composition over time. This gender preference is so strong that based on labor supply alone, most occupations have multiple equilibria, meaning that historical segregation could determine whether an occupation is male or female. However in the labor market, unlike in a classic Schelling model, compensating differentials moderate the impact of the gender preference, and ensure that all occupations have a unique equilibrium in fraction female. Therefore even at the high level of gender preference I estimate, manipulating the gender composition of an occupation alone has no long run impact on segregation.}

@@

Occupation gender segregation is a persistent feature of the labor market that continues to contribute substantially to the gender wage gap, especially as the education gap between men and women has closed and even reversed \cite{Blau2006, Blau2013, Kahn2016a}. In recent literature, many factors have been identified as contributing causes of segregation, including preferences over amenities (\citeA{Olivieri2014a}, \citeA{Wiswall2013})\footnote{For overview see \citeA{ Bertrand2011} and \citeA{Cortes2017a}.}, productivity and skill differences (\citeA{Baker2015}), and discrimination by occupation (\citeA{Hsieh2016}).

Recent research using job satisfaction data suggests that women prefer not to work in male occupations \cite{Usui2008, Lordan2015}. Such a preference could operate at the occupation level, through gender identity effects \cite{Akerlof2000}, or through the work environment itself. Since the fraction female in an occupation evolves endogenously according to worker choices, preferences over the fraction female can produce outcomes such as multiple equilibria and ``tipping" at the occupation level. \citeA{Pan2010} documented ``tipping" patterns in U.S. occupations from 1940-1980 and posited a Schelling-style model of stable and unstable equilibria in the fraction female to explain the phenomenon. I illustrate this model in Appendix \ref{sec.Pan}.

%\footnote{Figure \ref{fig:tipping} in Appendix \ref{sec.Pan} illustrates the concept of ``tipping" as described in \citeA{Pan2010}.}
% \citeA{Pan2010} uses a regression discontinuity design to find thresholds after which occupations move rapidly from male to female,

%I use a structural model to disentangle these factors empirically using data on wages and observed matches. I then use the model to find equilibria in the gender composition, and predict the future patterns of segregation, responsiveness to shocks, and historical dependence.

%Preferences over occupation gender composition itself have received less empirical treatment. Such a preference could operate at the occupation level, through gender identity effects \cite{Akerlof2000}, or through the work environment itself. Using several surveys of job satisfaction, and employment transitions, \citeA{Lordan2015} find evidence consistent with a female preference for working in occupations with more women, and also some evidence for male preferences for working with men. Their estimates are robust even to controls for firm level fraction female and remain even after the inclusion of occupation task measures, suggestive of at least some preference over the gender composition at the occupation level.

% Black2010, Bronson2013, \cite{Goldin} bergmann
% Need a better citation for discrimination by occupation!!!!!!!

%The gender composition, and any correlated attributes, evolve endogenously from worker and firm choices. As a result, it is possible to observe ``tipping dynamics", or rapid movements in the gender composition, and multiple equilibria in the gender composition. \citeA{Pan2010} documents such patterns in the U.S. from 1940-1980, and posits a Schelling-style model of stable and unstable equilibria in the fraction female. Understanding the mechanisms of ``tipping" is important because some occupations may be highly responsive or non-responsive to policy levers intended to influence gender segregation. Outcomes will depend on the location and existence of equilibria in the gender composition. Finding the locations of these equilibria is difficult because they depend on the interaction of firm preferences, worker preferences, and equilibrium wages. I use a structural model to disentangle these factors empirically using data on wages and observed matches. I then use the model to find equilibria in the gender composition,  and predict the future patterns of segregation, responsiveness to shocks, and historical dependence.

Understanding the what has caused these ``tipping" patterns is important to understanding how gender segregation will respond to future labor market policies. In particular, if ``tipping" is caused by multiple equilibria in the gender composition, then some occupations may be highly responsive or non-responsive to policy levers, depending on the distance from stable equilibria and the location of tipping points. In addition, the existence of multiple equilibria would mean there could be other segregation patterns that are stable and even welfare enhancing that could be targeted by policies that manipulate the gender composition of occupations alone.

%I empirically disentangle mechanisms that affect the stability of segregation patterns by occupation. A
Finding the locations of equilibria in the fraction female is a difficult empirical problem. Equilibria depend on the interaction of firm preferences, worker preferences, and wages, which are also a function of firm and worker preferences. Therefore I estimate a structural model in order to predict past and future segregation patterns including multiple equilibria. Because segregation emerges from both sides of the market, and workers and firms are optimizing and connected through equilibrium wages, a matching model with observed transfers is a natural candidate to provide enough structure for this context. A one-sided model ignoring the firm side would not only bias estimates of worker preferences, it would also bias estimates of the location of tipping points and multiple equilibria. Although the model was developed specifically for segregation in the labor market, it could be used for estimation in any context where a price mechanism clears a two-sided market, with or without an in-group preference or endogenous amenity.

% The structural model makes it possible to simulate the outcome of a wide variety of shocks including job amenity requirements that affect non-wage utility, labor supply, and labor demand.

%As shown in simulations in the results section, firm choices have a large effect on tipping patterns. For example, in the case where women strongly prefer to work with women, firms can lower female wages as women enter their occupation due to the compensating differential. Lower female wages actually tend to prevent movement to 100\% female, as the lower wages limit the female labor supply without undercutting all potential male entrants (because wages cannot go negative and men are generally more valued). On the other hand, as women leave an occupation, firms are forced to raise female wages to compensate women for a more male environment. Depending on the extent to which firms are willing to pay more for men than women, female wages may go up enough that they are ``priced out" of the occupation, in that firms prefer to hire only men.

%It is also critical to have accurate estimates of worker preferences over the fixed attributes of each occupation, as well as the wage and fraction female, which are endogenous attributes. Men and women valuing occupations differently could be a major cause of the persistent segregation observed over time and is important to distinguish from sorting caused by preferences over fraction female. Sorting based on persistent non-wage attributes of occupations is more likely to be optimal in that it may be costly to change these attributes.

%Unlike the two-sided model that I propose,

%A one-sided model of worker choice is not sufficient because worker-job matches and wages are selected through the optimal choices of both firms and workers. Therefore ignoring the firm side would not only bias estimates of worker preferences, but also potentially affect the location of tipping thresholds and multiple equilibria.

%Since both workers and firms care about wages, I must also model how market clearing wages are determined as a function of worker and firm preferences.

%To separate the firm's choice from the worker's choice I must first model how the wages are determined to clear the market.

%A model of how equilibrium wages are determined is needed to separate out the impact of the firm's choice versus the worker's choice. The worker's choice depends on the fixed non-wage attributes of the occupation, which may be differentially valued by gender, as well as the wage and fraction female. Meanwhile the wage that the worker faces is determined in equilibrium including both firm and worker preferences.

% Changes over time in the fraction female are the result of optimal choices on both sides of the market.

% \begin{align*}
% \text{\textit{Share of gender in occupation}} =& f(\text{\textit{Non-wage attributes}},\text{\textit{ Fraction female}}, \text{\textit{\textbf{reservation wages}}}) \\
%\text{\textit{\textbf{reservation wages}}} =& g(\text{\textit{firm preferences, worker preferences, number of worker and jobs}})
% \end{align*}

%The fraction female is not endogenous within a given cohort of workers. Instead new cohorts take the decisions of previous cohorts as given.

For tractability I treat occupation choice as a static choice. Workers choose occupation once at the beginning of their career, based on contemporary characteristics of the occupations, and do not anticipate future changes in the labor market. In addition I assume that workers do not anticipate the simultaneous decisions of their peers, but rather respond to the existing gender composition. Relaxing this assumption, or allowing workers to myopically choose occupation more often during their lifetime, would lead to faster convergence to equilibrium in the fraction female, but would not change the overall dynamics of the model. Identification also relies on a large market assumption, meaning a continuum of workers and jobs of each type, and no unobserved individual worker characteristics valued by firms. Unobserved productivity of individual workers could be included in theory, but would result in very weak identification and likely computational intractability.\footnote{The information afforded from the wage distribution, which is key to identifying the model, would become a function of two distributions of unobserved heterogeneity affected by simultaneous selection problems from both sides of the market. So whether an observed wage results from an unobservably talented worker or and unobservably wonderful job, or a combination of both, would depend on all parameters of the model and integration over both distributions of unobserved heterogeneity.} Identification assumptions will be discussed in more detail below.

%To identify the model I assume that workers make lifetime occupation choices early in life, and do not anticipate changes in the fraction female later in life that would induce them to switch occupations. The expectation of workers is that the fraction female they observe among older workers when they choose their occupation is a good approximation for the fraction female they will face during their lifetime. Therefore the fraction female is not endogenous within a given cohort of workers, rather it changes endogenously across cohorts.

I take a two-step estimation approach. First I estimate the locations of the reservation wage distributions of workers by occupation and gender, and labor demand for male and female workers by occupations. This is done with maximum likelihood estimation on the observed wage distribution, assuming wages are the result of transferable utility matching, to be discussed below. Second, I estimate worker labor supply, taking the locations of the reservation wage distributions as given. I disentangle the effect of wages, fraction female, and other job attributes using a shift-share instrumental variables strategy. I instrument for fraction female and reservation wage at the occupation level by exploiting Bartik-style variation in exposure to industries and male and female labor force participation, as well as the estimated firm-side preferences from the first stage.

In the first stage of estimation, I find that firms are often willing to pay much more to hire male than female workers, but the degree to which this is true varies greatly across occupation. This gap may reflect discrimination, differences in labor force attachment over the lifetime, skill differences, or job sorting not captured in my occupation categories. The gap in the willingness to pay for female vs. male workers is lower in female dominated occupations. This might be because women are selecting into occupations in which they are more valued or have comparative advantage, or because the way work is valued might vary with the fraction female in the occupation \cite{Levanon2009, Harris2018}, or some combination of both.

@@

In the second stage of estimation I find that women care strongly about the number of women in an occupation, but no evidence that men care about the number of women. If either men or women have preferences over the fraction female of occupations then we could observe multiple equilibria and ``tipping" patterns that result from moving between these equilibria. Despite the relatively strong in-group preference on the part of women, I find that equilibrium wages are able to adjust to prevent the emergence of multiple equilibria, and therefore tipping between equilibria. A stronger preference would be required to produce these phenomena.

% going from 0% to 100% female is about equivalent to an extra 6 million dollars in lifetime income
% and yet I still don't see multiple equilibria????

Women's preference for not working in male occupations still affects the dynamics of occupation segregation despite not being strong enough to produce multiple equilibria. In particular the preference accentuates the impact of gendered changes to occupation quality and demand, which can lead to rapid movement in the fraction female. An illustrative case study is insurance adjustors, which is an occupation observed to have tipped from male to female by \citeA{Pan2010}. An increase in the willingness-to-pay for female workers combined with the preference of women for working with women led to the observed ``tipping".  I predict that if it were not for women preferring to work with women, insurance adjusters would have become only 50\% female, rather than the observed over 70\% female (from a starting point of 20\% female in 1960), but this rapid movement was \textit{not} the result of moving between different stable equilibria.

@@

The gender preference also leads to more segregated outcomes in the cross-section. With no gender preference, the model predicts no occupations with fraction female greater than 70\% or less than 10\%, and a Duncan segregation index of 24\% in 2012, meaning 24\% of male or female workers would have to change occupations to make all occupations 50\% female. With the gender preference, the Duncan index is predicted to be almost twice as high at 47\%.\footnote{The Duncan index in my observed 2012 data is 41\%.}

@@

Although the gender preference I estimate is very strong, it is largely mitigated by compensating differentials. In a model without endogenous wages, all occupations eventually converge to either 0\% or 100\% female, and most occupations have multiple equilibria in the fraction female. Wage adjustment leads to more moderate outcomes. As women enter an occupation, firms can hire the same number of women at lower cost, which dampens the increase in female labor supply. Likewise as women exit an occupation, firms must pay the remaining women more, which slows the exit of women.

@@

%and as women leave an occupation higher and higher wages are needed to retain them, it is possible for women to be completely priced out of an occupation as it becomes more male. On the other hand, it is hard for men to be priced out of a female occupation, given that firms are willing to pay more for men. As more women enter an occupation women's wages go down as the occupation becomes more attractive, but wages cannot go negative, so firms continue to hire some men into every occupation.

There are two key policy implications of the estimated model. First is that that current sorting patterns are surprisingly resilient to short-run shocks. Policies that only temporarily change the attractiveness of occupations to men vs. women have no long run impact on the equilibrium fraction female. Second, sorting patterns are not historically dependent. For example, if all occupations were 50\% female in 1960, gender segregation would still be very similar to what it is today. So policies to reduce segregation today must focus on the fundamentals of occupation amenities and labor demand rather than short run shocks to overcome historical inertia. But changes in labor demand and supply can still move the existing equilibrium, and such movements are reinforced by women wanting to work with women.

@@

Section \ref{literature} provides a brief overview of the literature. Section \ref{model} introduces the structure of the transferable utility matching model. Section \ref{equilibrium} describes equilibrium wages. Section \ref{empirical} describes the empirical specification and first stage of estimation, recovering the firm side parameters and reservation wages. Section \ref{stage2} describes the second stage of estimation, the decomposition of worker utility taking reservation wages as given. Section \ref{data} describes the data sources. Section \ref{results} presents results, and Section \ref{counterfactuals} presents simulations of segregation dynamics, including counterfactuals.

@@

\section{Literature Review} \label{literature}

A number of papers have studied worker preferences for tasks and amenities by occupation. \citeA{asadullah2008} and \citeA{Reed1994} find that men and women have similar preferences over amenities using job satisfaction surveys and job transitions respectively. On the other hand, \citeA{Goldin2011} and \citeA{Goldin2014} find higher demand from female workers for flexibility and shorter hours to be important in explaining sorting and wage differentials. I find large variation in male and female preferences for different occupations, which could include any occupation amenities that are fixed over time, such as flexibility, task requirements,\footnote{There is evidence that changes in task requirements over time is important in explaining gender gaps \cite{Black2010, atalay2017}. In a robustness check I control for the time-varying task measures of \citeA{atalay2017} and find qualitatively similar results for gender preference over fraction female.} hours, or risk.

Several papers look for preference over the gender composition of the occupation. Using job satisfaction data from the NLSY, \citeA{Usui2008} finds that women dislike male occupations and receive a compensating differential to work in them. Similarly \citeA{Lordan2015} use job satisfaction and transition data to find that even controlling for occupation task measures and firm level fraction female, there is still evidence for women disliking male occupations.\footnote{The estimates in \citeA{Lordan2015} are robust even to controls for firm level fraction female and remain even after the inclusion of occupation task measures, suggestive of at least some preference over the gender composition at the occupation level.} Unlike this literature my paper explicitly models how such a preference factors into occupation choice, including the impact of compensating differentials on equilibrium wage levels.

The methodological contribution of my paper builds on a long legacy of literature on theoretical matching. \citeA{Shapley1972a} showed that with transferable utility the stable matching coincides with the competitive equilibrium. \citeA{Crawford1981} and \citeA{RothSotomayor} further developed mechanisms for convergence to the competitive equilibrium, and conditions for the uniqueness of the prices, or wages, that support it. More recently \citeA{Choo2006} pioneered empirical estimation of transferable utility models with an application to the marriage market. \citeA{Salanie2014a} and \citeA{Dupuy2017} have presented the theoretical possibility of using observed transfers for identification. My paper is the first to my knowledge to take this insight to an empirical application, using wages to disentangle firm and worker preferences in the labor market.

A possible consequence of workers having preferences over groups is ``tipping", or rapid movements in segregation patterns. The literature on tipping includes the seminal work of \citeA{Card2008}, identifying tipping points for racial segregation in neighborhoods, \citeA{Caetano2017} finding tipping and multiple equilibria in the racial composition of schools, and of course \citeA{Pan2010} using regression discontinuity to find tipping points in fraction female in occupations. I focus on disentangling the mechanisms underlying tipping, and am the first to my knowledge to examine the role of equilibrium wages or prices in such a market.
